---
title: "Introduction to the `learn2count` package"
author: "Kim Hue Nguyen and Davide Risso"
date: "Last modified: November 7, 2020; Compiled: `r format(Sys.time(), '%B %d, %Y')`"
bibliography: refs.bib
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteEncoding{UTF-8}
---

# Installation

The recommended way to install the `learn2count` package is

```{r, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("learn2count")
```


# Introduction

```{r options, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(warning=FALSE, error=FALSE, message=FALSE)
set.seed(123)
```

This vignette provides an introductory example on how to work with the `learn2count` package, which implements the structure learning methods proposed in @nguyen2020structure. First, let us load the packages and set serial computations.

```{r load_packs}
library(foreach)
library(learn2count)

# Register sequential computations 
# see the `doParallel` package for parallel computations
registerDoSEQ()
```

# The ZINB model

ZINB is a general and flexible model for Structure learning of high-dimensional zero-inflated count data, such as those recorded in single-cell RNA-seq assays. Given \(n\) samples (typically, \(n\) single cells) and \(p\) features (typically, \(p\) genes) that can be counted for each sample, we denote with \(X_{is}\) the count of feature \(s\) (\(s=1,\ldots,p\)) for sample \(i\) (\(i=1,\ldots,n\)). To account for various technical and
biological effects, typical of single-cell sequencing
technologies, we model \(X_{is}\) conditional to all possible subsets  of variables \(\mathbf{x}^{(i)}_{K}, \, K\subseteq V\) is a zero-inflated negative binomial (zinb) distribution with parameters \(\mu_{is|K}\), \(\theta_{s|K}\), and
\(\pi_{is|K}\), and consider the following regression models for the parameters:

\begin{equation}\label{localcondzinb}
    f_{zinb}(x_{is};\mu_{is|K},\theta_{s},\pi_{is|K}|\mathbf{x}^{(i)}_{K\setminus\{s\}})= \pi_{is|K}\delta_0(x_{is})+(1-\pi_{is|K})f_{nb}(x_{is},\mu_{is|K},\theta_{s}|\mathbf{x}^{(i)}_{K\setminus\{s\}}),
\end{equation}
where \(\delta_0(.)\) is the Dirac function,   \(\pi_{is|K}\in [0,1]\) is the probability that a 0 is sampled from a distribution degenerate at zero and \(f_{nb}(.,\mu,\theta)\) denotes the probability mass function of the negative binomial (NB) distribution with mean $\mu$ and inverse dispersion parameter \(\theta.\) We assume that
\begin{align}
	\ln(\mu_{is|K})&= \nu^{\mu}_{s|K}+\sum_{t\in K\setminus\{s\}} \beta^{\mu}_{st|K}x_{it}\label{link-mu},\\
\text{logit}(1-\pi_{is|K})&=\nu^{\pi}_{s|K}+\sum_{t\in K\setminus\{s\}} \beta^\pi_{st|K}x_{it}.\label{link-pi}
\end{align}

When both  $\mu_{is|K}$ and $\pi_{is|K}$,  are not constant, a missing edge between node $s$ and node $t$ corresponds to the condition $\beta^{\mu}_{st|K}=\beta^{\mu}_{ts|K}=\beta^{\pi}_{st|K}=\beta^{\pi}_{ts|K}= 0, \,\, \forall K \subseteq V\setminus \{s\}.$ On the other hand, one edge between node $s$ and node $t$ implies that at least one of four parameters $\beta^{\mu}_{st|K}, \beta^{\mu}_{ts|K}, \beta^{\pi}_{st|K}, \beta^{\pi}_{ts|K}$ is different from 0.

This specification defines a family of models that includes the most common models employed  with count data and embraces a variety of situations. It is quite evident that, when 
$\pi_{is|K} = 0, \,\, \forall \ K \subseteq V\setminus \{s\},$ the model reduces to a NB distribution, which, in turn reduces to a Poisson distribution when the inverse dispersion parameter  $\theta_s$ tends to infinity. When $\pi_{is|K} > 0,$ zero-inflation comes into play and zero-inflated Poisson and NB models can be considered. In this case, when $\beta^{\pi}_{st|K} = 0,\,\, \forall \,\,t\in K\setminus\{s\},$  the neighborhood of a node $s$ is defined to be the set of effective predictors of $\mu_{is|K}$ and consists of all nodes $t$ for which $\beta^{\mu}_{st|K} \ne 0.$ On the other side, when $\beta^{\mu}_{st|K} = 0,\,\, \forall\,\, t\in K\setminus\{s\},$  the neighborhood of a node $s$ is defined to be the set of effective predictors of $\pi_{is|K}$ and consists of all nodes $t$ for which $\beta^{\pi}_{st|K} \ne 0.$ In other words, the family includes models in which the structure of the graph is attributable only to one of the two parameter components, $\pi_{is|K}$ or $\mu_{is|K}.$ 

# Simulated dataset

To illustrate the main functionalities of the package, we first make use of simulated data. This also allows us to showcase the simulation function that we provide as part of `learn2count`.

First, we use the `gRbase` package to simulate a graph with 10 nodes.

```{r set_graph}
library(gRbase)
graph <- ug(~ 1:2:3 + 3:4:5 + 4:5:6 + 6:7 + 8:5:9:10)
plot(graph)
adj <- as(graph, "matrix")
```

We can now use this adjacency matrix to simulate some data, for instance with `n=100` observations.

```{r sim}
set.seed(1130)
mat <- simdata(n=200, p=10, B=adj, family="Poisson", mu=5, mu_noise=1)
mat[1:3,1:3]
```

Finally, we can use the main `PCzinb` function to estimate the graph.

```{r est_sim}
est <- PCzinb(mat, method="poi", maxcard=3, alpha=0.1)
sparse_est <- as(est, "dgCMatrix")
graph_est <- as(sparse_est, "graphNEL")
plot(graph_est)
# sum(abs(est-adj)[lower.tri(est)])
```

# Mouse OE data

To illustrate the typical use of the package on real data, we analyze a set of cells, whose RNA profiles are assayed after injury of the mouse olfactory epithelium (OE), to characterize Horizontal Basal Cells (HBC) and their descendants during regeneration [@Fletcher2017], available through the `scRNAseq` package.

```{r datain}
library(scRNAseq)
library(scran)
sce <- FletcherOlfactoryData()
```

We focus here on the neuronal lineage, i.e., the cells that from HBC mature into neurons.
Furthermore, we select the set of 100 most variable genes and the 200 cells with the highest means.

We use this low-dimensional dataset in this vignette for computational reasons, but a real use case would involved a greater number of genes and cells, see @nguyen2020structure for a more realistic example.

However, it is in general a good idea to focus on a subset of highly-variable genes, in order to remove transcriptional noise and focus on the more biologically meaningful signals.

```{r fletcher}
neuronal <- sce[,!sce$cluster_label %in% c("iSus", "mSus", "MVC1", "MVC2")]
var <- modelGeneVarByPoisson(neuronal)
var <- var[order(var$bio, decreasing = TRUE),]
top100 <- neuronal[rownames(var)[1:100],]

means <- colMeans(counts(top100))
names(means) <- colnames(top100)
means <- sort(means, decreasing = TRUE)
top <- top100[,names(means)[1:200]]
top
```

## Preprocessing

We now apply a normalization procedure that allows us to get numerically more stable and computationally more efficient results.

While we show in @nguyen2020structure that the results with and without preprocessing are similar, we generally recommend to apply this preprocessing to real data, as it can speed up the computation substantially.

Our preprocessing strategy is described in detail in @nguyen2020structure. Briefly, we match the 95 percentiles across cells to account for differences in sequencing depth and we adjust the data to be closer to a zinb distribution by using a power transformation $X^\alpha$,  where $\alpha \in [0,1]$ is chosen to minimize the distance between the empirical distribution and the zinb distribution, measured by Kolmogorov-Smirnov statistics.

This procedure is implemented in the `QPtransform()` function.

```{r preprocess}
top <- QPtransform(top)
```

## PCzinb

We are now ready to reconstruct the graph using our PCzinb algorithm. The function, when applied to a SummarizedExperiment (or object inheriting from it) returns a SummarizedExperiment with the adjacency matrix of the graph in the metadata slot.

```{r zinb}
# top <- PCzinb(top, method="zinb1")
adj <- PCzinb(t(counts(top)), method="zinb1")
```

`method="zinb1"` specifies a model in which the structure of the graph is attributable to both of the two parameter components $\mu$ and $\pi$ of the zero-inflated negative binomial. 

If we want to infer the structure from only $\mu$, then we may specify `method="zinb0"`.

Alternatively, one can set `method="nb"` to fit negative binomial models, or `method="poi"` to fit Poisson models. While less flexible, these method are much more computataionally efficient.

## Parameter choice

The parameter `maxcard` controls the maximum number of varibales that is considered in the conditional sets. In this case, `maxcard` is set to its default value of 2. 

The parameter `alpha` is the significant level of the tests. By default it is set to $\alpha=2(1-\Phi(n^{0.2}))$, which is chosen to control the Type I error of all tests [@JMLR:v22:18-401].


Finally, the parameter `extend=TRUE` considers the union of the tests to infer the edge of the graph. If it is set to `FALSE` it will consider the intersection instead.

# References

